ODGI: understanding pangenome graphs

### 2021-10-04 manuscript/main.tex
metrics: figure, bib, script

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}

@@

\title[short Title]{ODGI: understanding pangenome graphs}
\author[Heumos, \textit{et~al}.]{
Andrea Guarracino\,$^{\text{\sfb 1} \dagger}$,
Simon Heumos\,$^{\text{\sfb 2} \dagger}$,
Sven Nahnsen\,$^{\text{\sfb 2}}$,
Pjotr Prins\,$^{\text{\sfb 3}}$,
and Erik Garrison\,$^{\text{\sfb 3}*}$
}

@@

% sort / layout / NO tension! We keep that for the sorting paper
Allows us to understand the sparse structures typically found in pangenome graphs.

@@

Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs. ODGI also offers more advanced tools for the interrogation of the graphs. \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to collapsed sequences with highly identical repeats. Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation.

@@

@@

Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.

>>>>

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}
\usepackage{subcaption}

@@

\title[short Title]{ODGI: understanding pangenome graphs}
\author[Guarracino, Heumos \textit{et~al}.]{
Andrea Guarracino\,$^{\text{\sfb 1} \dagger}$,
Simon Heumos\,$^{\text{\sfb 2} \dagger}$,
Sven Nahnsen\,$^{\text{\sfb 2}}$,
Pjotr Prins\,$^{\text{\sfb 3}}$,
and Erik Garrison\,$^{\text{\sfb 3}*}$
}

@@

% sort / layout / NO tension! We keep that for the sorting paper

% Allows us to understand the sparse structures typically found in pangenome graphs.

Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses. Uncovering their latent topology allows us to understand the sparse structures typically found in pangenome graphs.

\texttt{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension. Nodes can be sorted randomly, topologically, by breaking cycles in the graph, by grooming, or by using the novel path-guided stochastic gradient descent (PG-SGD) algorithm.  can be combined together to a series of sorts.

@@

Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length. MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs. ODGI also offers more advanced tools for the interrogation of the graphs. \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs. This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph. Collapsed sequences with highly identical repeats indicate intricate loci. Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation. In Fig. \ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.

\include{fig_metrics.tex}

@@

rdf sparql side, bring in annotation for great biological interpretation

some cancer stuff?!

(recent) interactive visualization tools?!

@@

Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
TODO!
